---
title: "final plot"
format: html
editor: visual
---

## Introduction

## Explanatory Analysis

### Year vs prevalence

```{r}
#| echo: false
#| warning: false
#| message: false
library(ggplot2)
library(dplyr)
library(kableExtra)
library(sjPlot)
library(performance)
library(DHARMa)
library(patchwork)
obesity<-read.csv("DAProject5.csv")
obesity$Year<-as.factor((obesity$Year))
obesity$Obese<-as.factor((obesity$Obese))
obesity$Education<-as.factor((obesity$Education))
obesity$Veg<-as.factor((obesity$Veg))
obesity$Fruit<-as.factor((obesity$Fruit))
obesity$Sex<-as.factor(obesity$Sex)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: false

glimpse(obesity)

```

```{r}
#| echo: false
#| label: tbl-year
#| tbl-cap: Prevalence of obesity over 2013-2016

yearprop<-prop.table(table(obesity$Year,obesity$Obese),1)
yearprop<-as.data.frame(yearprop)
colnames(yearprop)<-c("Year","Obese","Prevalence")
yearprop100<-prop.table(table(obesity$Year,obesity$Obese),1)*100
yearprop100<-as.data.frame(yearprop100)
colnames(yearprop100)<-c("Year","Obese","Prevalence(%)")
yearprev<-yearprop100 %>%
  dplyr::filter(Obese=="Yes")
yearprev<-yearprev[,-2]
kable(yearprev,
      digit=3)

```

```{r}
#| echo: false
#| message: false
#| warning: false 
#| label: fig-year
#| fig-cap: Prevalence of obesity across 2013-2016

ggplot(yearprop, aes(x = Year, y = Prevalence, fill = Obese)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent,
                      limits = c(0,1)) +
  labs(x="Year", y="Prevalence",
       title="Obesity Prevalence by Year")  

```

### Sex vs prevalence

```{r}
#| echo: false
#| label: tbl-sex
#| tbl-cap: Prevalence of obesity by sex
sexprop<-prop.table(table(obesity$Sex, obesity$Obese),1)
sexprop<-as.data.frame(sexprop)
colnames(sexprop)<-c("Sex","Obese","Prevalence")
sexprop100<-prop.table(table(obesity$Sex, obesity$Obese),1)*100
sexprop100<-as.data.frame(sexprop100)
colnames(sexprop100)<-c("Sex","Obese","Prevalence(%)")

sexprev<-sexprop100 %>%
  dplyr::filter(Obese=="Yes")
sexprev<-sexprev[,-2]
kable(sexprev,
      digit=3)
```

```{r}
#| echo: false
#| label: fig-sex
#| fig-cap: Obesity prevalence by sex
ggplot(sexprop, aes(x = Sex, y = Prevalence, fill = Obese)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent,
                     limits = c(0,1)) +
  labs(
    title = "Obesity Prevalence by Sex",
    x = "Sex",
    y = "Prevalence"
  )
```

### Age vs prevalence

```{r}
#| echo: false
#| label: fig-ageboxplot
#| fig-cap: Age Distribution by Obesity Status
ggplot(obesity, aes(x = Obese, y = Age,fill=Obese)) +
  geom_boxplot() +
  labs(title="Age Distribution by Obesity Status")

```

### Education vs prevalence

```{r}
#| echo: false
#| label: tbl-edu
#| tbl-cap: Prevalence of obesity by education level
eduprop<-prop.table(table(obesity$Education, obesity$Obese),1)
eduprop<-as.data.frame(eduprop)
colnames(eduprop)<-c("Education","Obese","Prevalence")
eduprop100<-prop.table(table(obesity$Education, obesity$Obese),1)*100
eduprop100<-as.data.frame(eduprop100)
colnames(eduprop100)<-c("Education","Obese","Prevalence(%)")

eduprev<-eduprop100 %>%
  dplyr::filter(Obese=="Yes")
eduprev<-eduprev[,-2]
kable(eduprev,
      digit=3)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-edu
#| fig-cap: Obesity Prevalence by Education Level

ggplot(eduprop, aes(x = Education, y = Prevalence, fill = Obese)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent,
                     limits = c(0,1)) +
  labs(
    title = "Obesity Prevalence by Education Level",
    x = "Education Level",
    y = "Prevalence"
  )+
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```

### Vegetable intake

```{r}
#| echo: false
#| label: tbl-veg
#| tbl-cap: Prevalence of obesity by Vegetable intake
vegprop<-prop.table(table(obesity$Veg, obesity$Obese),1)
vegprop<-as.data.frame(vegprop)
colnames(vegprop)<-c("Vegetable_intake","Obese","Prevalence")
vegprop100<-prop.table(table(obesity$Veg, obesity$Obese),1)*100
vegprop100<-as.data.frame(vegprop100)
colnames(vegprop100)<-c("Vegetable_intake","Obese","Prevalence(%)")

vegprev<-vegprop100 %>%
  dplyr::filter(Obese=="Yes")
vegprev<-vegprev[,-2]
kable(vegprev,
      digit=3)
```

```{r}
#| echo: false
#| label: fig-veg
#| fig-cap: Obesity Prevalence by Vegatable intake
ggplot(vegprop, aes(x = Vegetable_intake , y = Prevalence, fill = Obese)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent,
                     limits = c(0,1)) +
  labs(
    title = "Obesity Prevalence by Vegetable Intake",
    x = "Vegetable Intake",
    y = "Prevalence"
  )

```

### Fruits vs prevalence

```{r}
#| echo: false
#| label: tbl-fruits
#| tbl-cap: Prevalence of obesity by Fruit intake
frprop<-prop.table(table(obesity$Fruit, obesity$Obese),1)
frprop<-as.data.frame(frprop)
colnames(frprop)<-c("Fruit_intake","Obese","Prevalence")
frprop100<-prop.table(table(obesity$Fruit, obesity$Obese),1)*100
frprop100<-as.data.frame(frprop100)
colnames(frprop100)<-c("Fruit_intake","Obese","Prevalence(%)")

frprev<-frprop100 %>%
  dplyr::filter(Obese=="Yes")
frprev<-frprev[,-2]
kable(frprev,
      digit=3)
```

```{r}
#| echo: false
ggplot(frprop, aes(x = Fruit_intake, y = Prevalence, fill = Obese)) +
  geom_col(position = "dodge")  +
  scale_y_continuous(labels = scales::percent,
                     limits = c(0,1))+
  labs(
    title = "Obesity Prevalence by Fruit intake",
    x = "Fruits Intake",
    y = "Prevalence"
  )
```

## Formal Analysis

### Year vs prevalence

$$logit(p_i)=\beta_0+\beta_1(Year)$$ {#eq-year}

where $p_i$ is the probability that an individual is obese.

```{r}
#| echo: false
#| message: false
#| warning: false 
#| label: tbl-yearmodel
#| tbl-cap: Coefficient of the logistic regression model
model_year <- glm(Obese ~ Year,
                  data = obesity,
                  family = binomial)

summary(model_year)

tab_model(model_year,transform=NULL)
```

Each coefficient compares that year with 2013(reference level)

Since p \> 0.05, there is no significant difference in obesity prevalence between years.

```{r}
#| echo: false
exp(coef(model_year))
```

odds are very similar across years

### Sex vs prevalence

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-sexmodel
#| tbl-cap: Coefficient of the logistic regression model 
model_sex <- glm(Obese ~ Sex,
                 data = obesity,
                 family = binomial)

summary(model_sex)

tab_model(model_sex,transform=NULL)
```

Since p=0.00246\<0.05, obesity prevalence differ by sex

```{r}
#| echo: false
exp(coef(model_sex))
```

Odds ratio \< 1 → males have lower odds of obesity compared with females.

### Age vs prevalence

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-agemodel
#| tbl-cap: Coefficient of the logistic regression model
model_age <- glm(Obese~ Age,
                 data = obesity,
                 family = binomial)

summary(model_age)

tab_model(model_age,transform=NULL) 
```

Since p-value\<0.05-\> age is a signicant factor in obesity prevalence

```{r}
#| echo: false
exp(coef(model_age))
```

Each 1-year increase in age increases odds of obesity by about 1.02%.

```{r}
#| echo: false
#| label: fig-age
#| fig-cap: Predicted relationshio between age and obesity prevalence
obesity$Obese_num <- ifelse(obesity$Obese == "Yes", 1, 0)
ggplot(obesity, aes(x = Age, y = Obese_num)) +
  geom_jitter(height = 0.05, alpha = 0.2) +
  stat_smooth(method = "glm", method.args = list(family = "binomial")) +
  labs(
    y = "Probability of Obesity",
    title = "Relationship Between Age and Obesity Prevalence" 
  )
```

### Education vs prevalence

```{r}
#| echo: false
#| label: tbl-edumodel
#| tbl-cap: Coefficient of the logistic regression model
#| warning: false
#| message: false

 model_edu <- glm(Obese ~ Education,
                 data = obesity,
                 family = binomial)

summary(model_edu)

tab_model(model_edu,transform=NULL)
 

```

gives model formula

look at pvalues

```{r}
#| echo: false
exp(coef(model_edu))
#compare odds of being obese to referenc group 
```

### Vegetable intake vs prevalence

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-vegmodel
#| tbl-cap: Coefficient of the logistic regression model

model_veg <- glm(Obese ~ Veg,
                 data = obesity,
                 family = binomial)

summary(model_veg)

tab_model(model_veg,transform=NULL)
```

pvalue=0.00298\<0.05, significant

interpretation of estimates/odd ratios?

```{r}
#| echo: false
exp(coef(model_veg))

##odd ratio
```

odd ratios

Individuals meeting recommended vegetable intake have lower odds of obesity compared with those who do not.

### Fruit intake vs prevalence

```{r}
#| echo: false
#| label: tbl-fruitmodel
#| tbl-cap: Log odds of the logistic regression model
#| message: false
#| warning: false

model_fruit <- glm(Obese ~ Fruit,
                   data = obesity,
                   family = binomial)

summary(model_fruit)

tab_model(model_fruit,transform=NULL)

```

pavlue=0.839-\>factor is insignicant

```{r}
#| echo: false
exp(coef(model_fruit))
```

### Full model with all variables

```{r}
#| echo: false
#| message: false
#| warning: false
full_model <- glm(Obese ~ Age + Sex + Education + Veg + Fruit + Year,
                   data = obesity,
                   family = binomial)

```

look at p-value -\> can remove year and fruit from model

### Reduced model (without year and fruit)

```{r}
#| echo: false
#| message: false
#| warning: false

reduced_model<- glm(Obese ~ Age + Sex + Education + Veg,
                                 data = obesity,
                                 family = binomial)

```

### Interaction model

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-modelcomplogodd
#| tbl-cap: Model comparison
model_interaction <- glm(
  Obese ~ Age + Sex + Education + Veg + Fruit +
    Age:Sex +
    Age:Education +
    Veg:Fruit,
  data = obesity,
  family = binomial
  )
tab_model(full_model,reduced_model,model_interaction, transform = NULL,
          dv.labels = c("Full model","Reduced model","Interaction Model"),
          show.aic = T)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-aic
#| tbl-cap: Model AIC comparison


compare_performance(full_model,reduced_model,model_interaction,metrics = "AIC") %>% 
  kable()


```

The interaction model has the lowest aic-\> best fit model

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-anova
#| tbl-cap: Model likelihood ratio test

anova(full_model,reduced_model, model_interaction, test = "Chisq") %>% 
  kable()
```

since p value for interaction model \<0.05, it significantly improves model fit , therefore the interaction model is preferred. (same conclusion as aic test)

```{r}
#| echo: false
#| message: false
#| warning: false 

library(sjPlot)

p1<-plot_model(
  model_interaction,
  type = "pred",
  terms = c("Age"),
  axis.title = c("Age", "Predicted probability of obesity"),
  title = " "
)
```

```{r}
#| echo: false
#| message: false
#| warning: false 

library(sjPlot)

p2<-plot_model(
  model_interaction,
  type = "pred",
  terms = c("Sex"),
  axis.title = c("Gender", "Predicted probability of obesity"),
  title = " "
)
```

```{r}
#| echo: false
#| message: false
#| warning: false 

library(sjPlot)

p3<-plot_model(
  model_interaction,
  type = "pred",
  terms = c("Education"),
  axis.title = c("socio-economic status", "Predicted probability of obesity"),
  title = " "
)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| echo: false
#| message: false
#| warning: false 

p4<-plot_model(
  model_interaction,
  type = "pred",
  terms = c("Veg", "Fruit"),
  axis.title = c("Lifestyle", "Predicted probability of obesity"),
  title = " "
)
```

```{r}
#| echo: false
#| message: false
#| warning: false 
library(sjPlot)

p5<-plot_model(
  model_interaction,
  type = "pred",
  terms = c("Age", "Sex"),
  axis.title = c("Age and sex", "Predicted probability of obesity"),
  title = " "
)
```

```{r}
#| echo: false
#| message: false
#| warning: false
(p1 + p2 + p3 + p4 + p5) +
plot_layout(ncol = 3) &
theme(plot.margin = margin(3,3,3,3))
```

### Model check

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-check
#| fig-cap: Final model diagnostic plots
library(performance)
check_model(model_interaction, check = c("pp_check","binned_residuals","outliers","qq"))
```
