---
title: "Group_05_Analysis"
author: "*Group 5: Chloe Au Yeung, Xinyan Tian, Cameron Fraser, and Sarah Khan*"
date: "03/16/2026"
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    toc-expand: true
    number-sections: false
execute:
  echo: false
---

# Introduction {#sec-intro}

In this report, we will begin using logistic regression to model outcomes of interest that take one of two categorical values. Then, we will use model diagnostics to assess if this model is the most reliable and fits the data best. Then we will look at the `DAProject.csv` dataset which contains the socio-economic and lifestyle factors from the 2013 - 2016 Scottish Health Surveys.

## Data Overview {#sec-data-overview}

In this report, we will use the data stored in `DAProject5.csv` which has 7 variables. The data comes from a Scottish Health Surveys study from the 2013 - 2016 Scottish Health Surveys on 14,018 individuals to study socio-economic and lifestyle factors. `DAPprojects.csv` contains the following columns:

-   `Age` - Age of individual

-   `Sex` - Sex of individual (Male / Female)

-   `Education` - Highest educational qualification of individual

-   `Veg` - Consume recommended daily vegetable intake (Yes / No)

-   `Fruit` - Consume recommended daily fruit intake (Yes / No)

-   `Year` - Year of the Scottish Health Survey

-   `Obese` - Indicator of individuals obesity classification (Yes / No)

Since this dataset contains a binary categorical variable `Obese` - classified as `Yes`/`No` - logistic regression may be considered as an appropriate modelling approach. We aim to create a predicative model that can determine the probability of an individual being classified as obese using variables `Age`, `Sex`, `Education`, `Veg`, and `Fruit`. Furthermore, we will assess if obesity has increased or decreased across the given years (2013-2016).

The main questions of interest are:

-   Has the prevalence of obesity in Scotland changed over the given years of the Scottish Health Survey?

-   Are there any differences in obesity by age, gender, socio-economic status, or lifestyle factors?

```{r}
#| echo: false
#| warning: false
#| message: false
library(ggplot2)
library(dplyr)
library(kableExtra)
library(sjPlot)
library(performance)
library(DHARMa)
library(patchwork)
library(ROCR)
library(yardstick)
obesity<-read.csv("DAProject5.csv")
obesity$Year<-as.factor((obesity$Year))
obesity$Obese<-as.factor((obesity$Obese))
obesity$Education<-as.factor((obesity$Education))
obesity$Veg<-as.factor((obesity$Veg))
obesity$Fruit<-as.factor((obesity$Fruit))
obesity$Sex<-as.factor(obesity$Sex)
```

# Explanatory Data Analysis {#sec-explanatory-data-analysis}

## Data Overview {#sub-sec-data-overview}

```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: false

glimpse(obesity)

```

Logistic regression is used when the outcome of interest takes one of two categorical values (e.g. `Yes`/`No`). In this case, the response variable takes the value 1 (defined as success, with probability $p_i$) or 0 (defined as failure, with probability $1 - p_i$). And the response is given by $Y_i$ with Bernoulli distribution.

The corresponding model uses the logit link function: $$
g(p_i) = 
\log\left(\frac{p_i}{1 - p_i}\right)
= \log\left(\frac{\mathbb{P}(Y_i = 1 \mid \underline{x}_i)}{\mathbb{P}(Y_i = 0 \mid \underline{x}_i)}\right)
= \underline{x}_i^\top \underline{\beta}
$$ {#eq-log-odds}

We can refer to @eq-log-odds as the log-odds, since $\frac{p_i}{1 - p_i}$ represents the odds. Taking the inverse of the logit link function gives the predicted probability:

$$
\mathbb{P}(Y_i = 1 \mid \underline{x}_i)
= \frac{\exp(\underline{x}_i^\top \underline{\beta})}{1 + \exp(\underline{x}_i^\top \underline{\beta})}
$$ {#eq-odds-1}

and it follows that:

$$
\mathbb{P}(Y_i = 0 \mid \underline{x}_i)
= \frac{1}{1 + \exp(\underline{x}_i^\top \underline{\beta})}.
$$ {#eq-odds-0} \## Visualisations {#sub-sec-visualisations}

To get an initial impression of the data set, we will look at the binary categorical variable `Obese` as our explanatory variable.

### Year vs prevalence

```{r}
#| echo: false
#| label: tbl-1
#| tbl-cap: Prevalence of obesity over 2013-2016

yearprop<-prop.table(table(obesity$Year,obesity$Obese),1)
yearprop<-as.data.frame(yearprop)
colnames(yearprop)<-c("Year","Obese","Prevalence")
yearprop100<-prop.table(table(obesity$Year,obesity$Obese),1)*100
yearprop100<-as.data.frame(yearprop100)
colnames(yearprop100)<-c("Year","Obese","Prevalence(%)")
yearprev<-yearprop100 %>%
  dplyr::filter(Obese=="Yes")
yearprev<-yearprev[,-2]
kable(yearprev,
      digit=3)

```

We can look at a bar plot of the proportion of `Obese` by `Year` to observe if there are changes in prevalence across the years.

```{r}
#| echo: false
#| message: false
#| warning: false 
#| fig-width: 4
#| fig-height: 2
#| label: fig-year
#| fig-cap: Prevalence of obesity across 2013–2016

ggplot(yearprop, aes(x = Year, y = Prevalence, fill = Obese)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("steelblue1", "steelblue4")) +
  scale_y_continuous(labels = scales::percent,
                     limits = c(0,1)) + 
  guides(fill = "none") +
  labs(x = "Year", y = "Prevalence")

```

We can see in @fig-year that there was an approximate consistency in proportion of individuals in the `Yes` obesity group. From @tbl-1, the proportion of individuals in the `Yes` obesity group appears, on average, at 30.5%.

Later, we will fit a logistic regression model to assess whether obesity prevalence has changed across the years, by examining whether `Year` is a significant predictor of the odds of individuals in the `Yes` obesity group.

### Sex vs prevalence

```{r}
#| echo: false
#| label: tbl-sex
#| tbl-cap: Prevalence of obesity by sex
sexprop<-prop.table(table(obesity$Sex, obesity$Obese),1)
sexprop<-as.data.frame(sexprop)
colnames(sexprop)<-c("Sex","Obese","Prevalence")
sexprop100<-prop.table(table(obesity$Sex, obesity$Obese),1)*100
sexprop100<-as.data.frame(sexprop100)
colnames(sexprop100)<-c("Sex","Obese","Prevalence(%)")

sexprev<-sexprop100 %>%
  dplyr::filter(Obese=="Yes")
sexprev<-sexprev[,-2]
kable(sexprev,
      digit=3)
```

In @tbl-sex, we can see that females in the `Yes` obesity group tends to be slightly higher than that of males.

### Age vs prevalence

```{r}
#| echo: false
#| fig-width: 4
#| fig-height: 2
#| label: fig-age
#| fig-cap: Age Distribution by Obesity Status
ggplot(obesity, aes(x = Obese, y = Age,fill=Obese)) +
  scale_fill_manual(values = c("steelblue1", "steelblue4")) +
  geom_boxplot() +
  labs(title="Age Distribution by Obesity Status")

```

@fig-age shows the distribution of age by obesity status. Individuals classified as obese tends to have slightly higher median ages compared to non-obese individuals.

### Education vs prevalence

```{r}
#| echo: false
#| label: tbl-edu
#| tbl-cap: Prevalence of obesity by education level
eduprop<-prop.table(table(obesity$Education, obesity$Obese),1)
eduprop<-as.data.frame(eduprop)
colnames(eduprop)<-c("Education","Obese","Prevalence")
eduprop100<-prop.table(table(obesity$Education, obesity$Obese),1)*100
eduprop100<-as.data.frame(eduprop100)
colnames(eduprop100)<-c("Education","Obese","Prevalence(%)")

eduprev<-eduprop100 %>%
  dplyr::filter(Obese=="Yes")
eduprev<-eduprev[,-2]
kable(eduprev,
      digit=3)
```

From @tbl-edu, proportion of individuals in the `Yes` obesity group variably increases across education levels. For example, individuals in the `No Qualifications` group have a higher prevalence of obesity (38.2%), than those in the `Degree of higher` group (24.7%).

### Vegetable intake

```{r}
#| echo: false
#| label: tbl-veg
#| tbl-cap: Prevalence of obesity by Vegetable intake
vegprop<-prop.table(table(obesity$Veg, obesity$Obese),1)
vegprop<-as.data.frame(vegprop)
colnames(vegprop)<-c("Vegetable_intake","Obese","Prevalence")
vegprop100<-prop.table(table(obesity$Veg, obesity$Obese),1)*100
vegprop100<-as.data.frame(vegprop100)
colnames(vegprop100)<-c("Vegetable_intake","Obese","Prevalence(%)")

vegprev<-vegprop100 %>%
  dplyr::filter(Obese=="Yes")
vegprev<-vegprev[,-2]
kable(vegprev,
      digit=3)
```

### Fruits vs prevalence

```{r}
#| echo: false
#| label: tbl-fruits
#| tbl-cap: Prevalence of obesity by Fruit intake
frprop<-prop.table(table(obesity$Fruit, obesity$Obese),1)
frprop<-as.data.frame(frprop)
colnames(frprop)<-c("Fruit_intake","Obese","Prevalence")
frprop100<-prop.table(table(obesity$Fruit, obesity$Obese),1)*100
frprop100<-as.data.frame(frprop100)
colnames(frprop100)<-c("Fruit_intake","Obese","Prevalence(%)")

frprev<-frprop100 %>%
  dplyr::filter(Obese=="Yes")
frprev<-frprev[,-2]
kable(frprev,
      digit=3)
```

Additionally, @tbl-veg and @tbl-fruits displays similar results for nutritional consumption `Yes`/`No` with the vegetable group intake being approximately the same and the `No` fruit intake group having a larger proportion (35.6% vs 29.8%).

We will extend the logistic regression model to see whether `Age`, `Sex`, `Education`, `Veg`, and `Fruit` are significant predictors of the odds of an individual being part of the `Yes` obesity group.

# Formal Data Analysis {#sec-formal-data-analysis}

## Model Specification {#sub-sec-model-specification}

We will now fit a logistic model to investigate the relationship between obesity and explanatory variables. Let $Y_i = 1$ if individual *i* is obese and $Y_i = 0$ otherwise. We will assume the following, as a result:

$$Y_i \sim \text{Bernoulli}(p_i)$$

where, $p_i = \mathbb{P}(Y_i = 1 \mid \underline{x}_i)$. The model is given by:

$$logit(p_i)=\alpha_0 + \beta_1 \,\mathbb{I}_{(\text{Year}_i = 2014)}
+ \beta_2 \,\mathbb{I}_{(\text{Year}_i = 2015)}
+ \beta_3 \,\mathbb{I}_{(\text{Year}_i = 2016)}$$ {#eq-year-model}

where: 2013 is the reference year. The intercept $\alpha_0$ represents the log-odds of an individual being obese in the reference year 2013, and the coefficients $\beta_1$, $\beta_2$, $\beta_3$ represent the change in log-odds in each year compared to 2013.

To fit a logistic regression model, we will conduct variable selection using backwards elimination to balance the model fit. We will also use Bayesian Information Criterion (BIC) as the selection criterion to determine the best model for our dataset. The performance of the model we select will be assessed using the Receiver Operating Characteristic (ROC) Curve.

## Results {#sub-sec-results}

### Year vs prevalence

We will fit a logistic regression model using the logit link function, with `Obese` as the response and `Year` as the explanatory variable:

```{r}
#| echo: false
#| message: false
#| warning: false 
#| label: tbl-yearmodel
#| tbl-cap: Coefficient of the logistic regression model
model_year <- glm(Obese ~ Year,
                  data = obesity,
                  family = binomial)

tab_model(model_year,transform=NULL)
```

The coefficients for all years appear close to zero. Additionally, the confidence intervals across all years include zero and p \>\> 0.05. This provides no evidence that year is a statistically significant predictor of obesity, which suggests that obesity prevalence has not changed over time

We can visualise the probabilities using the plot_model() function as follows:

```{r}
#| echo: false
#| message: false
#| warning: false 
#| fig-width: 4
#| fig-height: 3
#| label: fig-log-odds
#| fig-cap: Coefficient of the Logistic Regression Model
plot_model(model_year, show.values = TRUE, transform = NULL,
           title = "Log-Odds", show.p = FALSE)
```

@fig-log-odds shows the estimated log-odds coefficients for each year relative to the reference category (2013) are all close to zero, and their confidence intervals include zero. This indicates that there is minimal difference in log-odds of obesity across years. Adding onto this, @tbl-yearmodel shows large p-values, which further provides evidence that year is not a statistically significant predictor of obesity. Comparing coefficients of the regression model with that of the reference level appears to be approximately the same.

We can now obtain probability predictions for each year using the `predict()` function as follows:

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 4
#| fig-height: 4
#| label: fig-prob-obesity
#| fig-cap: Prob. of Obesity by Year

nd <- tibble(Year = factor(c(2013, 2014, 2015, 2016),
                levels = levels(obesity$Year)))

p <- predict(model_year,
             newdata = nd,
             type = "link",
             se.fit = TRUE) |>
  as.data.frame() |>
  mutate(fit_prob = plogis(fit),
    ll = plogis(fit - 1.96 * se.fit),
    ul = plogis(fit + 1.96 * se.fit)) |>
  bind_cols(nd)

ggplot(p, aes(x = Year, y = fit_prob)) +
  geom_point(size = 0.8) +
  geom_errorbar(aes(ymin = ll, ymax = ul), width = 0.1) +
  labs(x = "Year",
       y = "Prob. of obesity",
       title = "Predicted Probability of Obesity by Year") +
  ylim(0, 1)
```

From @fig-prob-obesity, we can predict the probability of obesity for each year. The probabilities appear to be constant at approximately 30%, with the confidence intervals overlapping. This, again, indicates that there is no significant change in obesity prevalence over time.

### Full model with all variables

We can fit a full model including variables `Age`, `Sex`, `Education`, `Veg`, and `Fruit`, as shown below:

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-fullmodel
#| tbl-cap: Log odds of the logistic regression model with all variables
full_model <- glm(Obese ~ Age + Sex + Education + Veg + Fruit,
                   data = obesity,
                   family = binomial)

tab_model(full_model,transform = NULL,
          dv.labels = "Full model")

```

The result of fitting a full model is that all predictors appear to be statistically significant except `Fruit`, Our full model indicates that fruit intake is not a statistically significant predictor for obesity due to the large p-value (0.942). From this insignificant p-value, we can remove fruit intake from our model.

### Reduced model (with fruit intake removed)

A reduced model, with fruit intake removed, is now fitted:

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-modelcomp
#| tbl-cap: Model comparison
reduced_model<- glm(Obese ~ Age + Sex + Education + Veg,
                                 data = obesity,
                                 family = binomial)

tab_model(reduced_model,transform = NULL,
          dv.labels = "Full model")

```

It appears all the reduced model predictors are statistically significant, as expected. Next, we will compare the BIC of the reduced model again the full model to supplement this conclusion.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-aic
#| tbl-cap: Model BIC comparison


compare_performance(reduced_model,full_model,metrics = "BIC") %>% 
  kable()
```

We can see that a full model with \`Fruit\` has a higher BIC in comparison to the reduced model. As a result, the reduced model is better balanced.

In addition to this, we can conduct a likelihood ratio test:

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-anova
#| tbl-cap: Model likelihood ratio test
anova(reduced_model, full_model, test = "Chisq") %>% 
  kable()

```

A likelihood ratio test comparing the reduced and full models yields a p-value of 0.999. Since p value\>\>0.05, this indicates that there is no evidence that the additional predictor `Fruit` in the full model improve model fit. Therefore, the reduced model is preferred.

### Model check

We will now investigate diagnostic plots to assess how well the model fits the data. This way we can assess whether our assumptions are met:

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-check
#| fig-cap: Final model diagnostic plots
library(performance)
check_model(reduced_model, check = c("pp_check","binned_residuals","outliers","qq"))
```

The posterior predictive check (top-left) shows that most residuals lie within the expected error bounds, with no strong systematic pattern. This suggests there is no systematic discrepancies between real and simulated data.

The binned residual plot (top-right) shows that most of the residuals lie within the expected error bounds, with about 95% of the residuals to fall inside the error bounds.

The influential observations plot (bottom-left) indicates that all points on this plot lie within Cook's distance. Therefore, there is no evidence of influential observations that affect the model.

The uniformity of residuals (bottom-right) closely follows the theoretical line, suggesting that the model assumptions are appropriate and to which the observed values do not deviate from the model expectations using simulated residuals.

Finally, we will assess the performance of the model using the Receiver Operating Characteristic (ROC) Curve:

```{r}
#| echo: false
#| message: false
#| warning: false

pred_results <- obesity %>%
  mutate(.pred_1 = predict(reduced_model, type = "response"))

roc_curve(pred_results,
          truth = Obese,
          .pred_1,
          event_level = "second") %>%
  autoplot() +
  labs(title = "ROC Curve",
       x = "1 - Specificity",
       y = "Sensitivity")
```

```{r}
#| echo: false
#| warning: false
#| label: fig-roc
#| fig-cap: ROC Curve

auc_val <- yardstick::roc_auc(
  pred_results,
  truth = Obese,
  .pred_1,
  event_level = "second")$.estimate

auc_val <- round(auc_val, 2)
```

The ROC curve in @fig-roc indicates if the model was able to distinguish between obese and non-obese individuals. Since the curve appears to be close to the diagonal line, this indicates the model only performs slightly better than guessing at random. The area under the curve (AUC) is 0.59, which suggests a moderate-poor predictive performance. This indicates that, although several predictors are statistically significant, the effect they have are not strong.

# Discussion {#sec-discussion}

# Conclusion {#sec-conclusion}

In conclusion, there is no evidence to suggest that obesity prevalence has any significant chance over the years we considered in this report. Moreover, there was a difference in obesity by fruit, however, no difference was observed by age, gender, socio-economic status, or vegetable consumption.

# References {#sec-references}

# Appendix {#sec-appendix}

```{r}
#| echo: false
#| message: false
#| warning: false

plot_sex <- ggplot(sexprop %>% dplyr::filter(Obese == "Yes"),
                   aes(x = Sex, y = Prevalence)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "By Sex",
    x = "Sex",
    y = "Obesity prevalence (%)"
  ) +
  theme_minimal()

plot_edu <- ggplot(eduprop %>% dplyr::filter(Obese == "Yes"),
                   aes(x = Education, y = Prevalence)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "By Education",
    x = "Education",
    y = "Obesity prevalence (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

plot_veg <- ggplot(vegprop %>% dplyr::filter(Obese == "Yes"),
                   aes(x = Vegetable_intake, y = Prevalence)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "By Vegetable Intake",
    x = "Vegetable intake",
    y = "Obesity prevalence (%)"
  ) +
  theme_minimal()

plot_fruit <- ggplot(frprop %>% dplyr::filter(Obese == "Yes"),
                     aes(x = Fruit_intake, y = Prevalence)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "By Fruit Intake",
    x = "Fruit intake",
    y = "Obesity prevalence (%)"
  ) +
  theme_minimal()

((plot_sex + plot_edu) / (plot_veg + plot_fruit))

```

## R Studio Language {#sub-sec-language}
